% =============================================================================
% 06_DISCUSSION — Interpretation, Implications, Limitations, Agenda
% =============================================================================
% Page target: ~2 pages
% Structure: Summary → Implications → Limitations → Research Agenda
% =============================================================================

\section{Discussion}
\label{sec:discussion}
\addcontentsline{toc}{section}{Discussion}

This discussion interprets the synthesized findings presented in
Section~\ref{sec:results} through an integrated analytical lens, examines
their implications for research, policy, and clinical practice, acknowledges
the limitations of this review, and concludes with a forward-looking
research agenda identifying the most pressing unanswered questions at the
intersection of health AI evaluation and governance.

\subsection{Summary of Key Findings}
\label{sec:discussion_summary}

The three thematic findings presented in the results are analytically
interdependent rather than independent. The \textbf{ecological validity
gap} in evaluation methodology does not merely reflect a technical
measurement problem — it has direct consequences for governance. When AI
systems are validated under conditions that do not reflect real-world
clinical complexity, governance frameworks based on those validations
will be structurally misaligned with the phenomena they seek to govern.
Similarly, the \textbf{governance capacity deficit} documented in health
systems is not simply a matter of underinvestment — it is partly a
consequence of evaluation frameworks that generate overconfident
validation evidence (suggesting AI systems are more ready for deployment
than they are), which reduces the perceived urgency of building
pre-deployment governance infrastructure. Finally, the \textbf{pilot
trap} can be understood as the downstream consequence of these two
upstream failures: systems validated on inadequate evidence, governed
by frameworks that assume absent capacity, predictably fail to transition
to routine care.

This interdependency implies that improving health AI outcomes requires
simultaneous action across all three dimensions — better evaluation
methodology, governance capacity investment, and implementation support
— rather than sequential approaches that treat evaluation, governance,
and implementation as separate phases.

\subsection{Implications}
\label{sec:discussion_implications}

\subsubsection{Implications for AI Evaluation Research}

The reviewed literature makes clear that the field needs a paradigm shift
from single-site, retrospective evaluation to prospective, multicenter,
ecologically grounded validation. This requires: (i) embedding AI evaluation
within real clinical workflows rather than controlled experimental setups;
(ii) adopting longitudinal evaluation designs that capture algorithmic drift
and performance evolution over time; (iii) developing clinical harm-asymmetric
performance metrics that reflect the differential consequences of false-positive
and false-negative errors in specific clinical contexts; and (iv) mandating
subgroup performance reporting to detect and quantify algorithmic bias across
population strata. The TRAC-Pain protocol \citep{bailes2026trac} offers a
methodological template that merits wider adoption and adaptation.

\subsubsection{Implications for Regulatory and Policy Practice}

The regulatory implications cut in two directions. First, current regulatory
frameworks must evolve from one-time pre-market authorization toward
\textbf{lifecycle-spanning governance} that mandates continuous post-market
performance monitoring, requires disclosure of training data sources, and
establishes enforceable equity and bias mitigation standards
\citep{abulibdeh2025illusion, warraich2024fda}. Second, regulators must
recognize that governance capacity in health systems is not pre-given and
cannot be assumed — it must be actively built through implementation
support, not just prescribed through standards \citep{tian2026pilottrap}.
This suggests that regulatory frameworks should incorporate implementation
support mechanisms, including regulatory sandboxes, technical assistance
programs, and mandatory pre-deployment organizational readiness assessments.

\subsubsection{Implications for Health Systems and Clinical Practice}

For health systems, the reviewed evidence points to the need to invest
in organizational governance infrastructure as a precondition for AI
deployment, not as an afterthought. This includes: establishing clear
accountability structures for AI-assisted clinical decisions; building
cross-functional governance committees that bring together clinical,
technical, legal, and patient representatives; developing operational
protocols for model updates, performance monitoring, and incident
response; and investing in workforce development to build algorithmic
literacy and clinical-AI workflow integration skills at scale.

\subsection{Limitations}
\label{sec:discussion_limitations}

This review is subject to five interrelated limitations that qualify the confidence with which its findings should be interpreted.

\subsubsection{Corpus construction and potential circularity}
\label{sec:limitation_circularity}

The literature corpus was assembled using a semi-automated pipeline that queries databases using keyword sets organized around five thematic axes pre-specified by the authors. While this approach enabled systematic and reproducible coverage of a rapidly evolving literature, it introduces a structural risk: the keyword architecture that defines what enters the corpus partly reflects the conceptual assumptions the framework was designed to test. Papers that document the three gaps in terms not anticipated by the search vocabulary, or that do so implicitly rather than explicitly, may have been systematically under-retrieved. The five-iteration refinement documented in Section~\ref{sec:corpus_prep} mitigates but does not eliminate this risk. Readers should treat the empirical gap frequencies as indicative of a directional pattern rather than as population-level estimates.

\subsubsection{Corpus size and generalizability after filtering}
\label{sec:limitation_corpus_size}

The v3 strict relevance filter, applied to correct for significant off-topic contamination in earlier iterations, reduced the analytically coded corpus from 341 candidate papers to 60. While the filter criteria are documented and motivated, a corpus of $n=60$ limits the statistical power of co-occurrence and interaction analyses. The conditional probabilities reported in Section~\ref{sec:cooccurrence} and Table~\ref{tab:conditional} are based on small cell counts: the gap2$\rightarrow$gap3 pathway, for instance, is documented in only three papers. The directional cascade finding is therefore best understood as a pattern warranted by the available evidence and consistent with the qualitative narrative of the cornerstone papers, rather than as a robust quantitative result. Replication with a larger, independently assembled corpus is needed before strong causal claims can be made.

\subsubsection{Scope of the regulatory comparative analysis}
\label{sec:limitation_regulatory}

Section~\ref{sec:regulatory} evaluates four regulatory frameworks, the EU AI Act, the US FDA framework, the UK MHRA framework, and the WHO Governance Guidance, against five evaluative conditions derived from the three-gap framework. While the conditions themselves are grounded in the v3 corpus, the regulatory assessment draws substantially on documentary sources and policy knowledge that extend beyond the coded literature. This is methodologically appropriate for a literature review, the corpus identifies what the gaps are; external regulatory knowledge is required to assess how existing frameworks address them, but it means that the regulatory findings cannot claim the same empirical grounding as the gap-frequency and interaction analyses. Moreover, the regulatory landscape is evolving rapidly: the EU AI Act entered into force in August 2024 and its implementing acts and conformity standards remain in development, and the FDA's predetermined change control plan guidance was issued in 2023 and is subject to revision. Assessments of framework adequacy should therefore be read as reflective of the state of governance as of mid-2026 and will require updating as regulatory elaboration continues.

\subsubsection{Language and geographic bias}
\label{sec:limitation_bias}

The pipeline queries predominantly English-language bibliographic databases, and the abstract-level screening retained only English-language publications. This creates a systematic bias toward research conducted in or published from North American, European, and Anglophone Asian health systems, contexts that differ substantially from the majority of health systems in which clinical AI is being deployed. The 10.0\% ELSE rate identified in Section~\ref{sec:stance} is partly an artifact of this collection bias: the equity and inclusion literature on non-Western AI implementation is a substantially larger body of work than the pipeline captures, and its systematic under-representation means that the three-gap framework as currently specified may be less applicable to low- and middle-income health system contexts than the evidence base suggests. The proposed fourth gap, equity and inclusion across non-Western health systems, should be understood not merely as a conceptual extension but as a corrective to a structural limitation of the review's evidence base.

\subsubsection{Absence of formal inter-rater reliability assessment}
\label{sec:limitation_inter_rater}

The structured content analysis in Section~\ref{sec:empirical} was conducted using LLM subagents coding in parallel batches of twenty papers, with the V3 protocol applied consistently across all coders. While this approach is documented and reproducible, the coding matrix and evidence appendix are provided as supplementary materials, it was not validated against human annotators, and no inter-rater reliability coefficient was computed for the gap and interaction codings. The stance classification layer introduced in Section~\ref{sec:stance} was assigned using post-hoc keyword lists derived from the same fields used to generate the V3.1 codes, introducing a risk of circularity in the stance distribution. These methodological choices were driven by practical constraints rather than by principled design, and they constitute a genuine limitation on the robustness of the coding scheme. Future work should validate the coding protocol through human annotation of a stratified sample of the corpus, ideally with independent coders blinded to the gap-frequency hypotheses the framework generates.
