% =============================================================================
% 06_CONCLUSION — Concluding Synthesis
% =============================================================================
% Rewritten 2026-06-26. The conclusion does NOT restate findings (already
% in abstract, discussion, and thematic) but synthesizes their implications,
% acknowledges the review's limits, and points to concrete next steps.
% =============================================================================

\section{Conclusion}
\label{sec:conclusion}
\addcontentsline{toc}{section}{Conclusion}

Building on the empirical analysis reported in Section~\ref{sec:empirical}
and the implications developed in Section~\ref{sec:discussion}, this review
contributes a three-gap framework --- ecological validity, governance
capacity, and pilot trap --- and the empirical evidence that these deficits
are structurally linked rather than independent. The framework's principal
value is not taxonomic but diagnostic: it specifies where in the AI
lifecycle intervention is most tractable, and where current regulatory
practice is most misaligned with the phenomena it seeks to oversee.

Several limitations qualify these findings. The v3 corpus is small (n=60)
and English-dominant, leaving non-Western implementation contexts,
non-English regulatory traditions, and equity outcomes in
low-resource settings systematically under-represented. The coding relied
on LLM-assisted structured content analysis validated against the
qualitative narrative of the four cornerstone papers, but without
formal inter-rater reliability assessment against human annotators.
Finally, the regulatory landscape is moving rapidly --- the EU AI
Act's implementing acts and the FDA's Predetermined Change Control Plan
guidance remain in development --- so the regulatory gap assessment is
time-bound to mid-2026 and will require updating.

Three implications follow for the field's principal audiences. For
\textbf{health AI researchers}, the priority is to retire single-site
retrospective validation as the dominant evaluation design and adopt
prospective, multicenter, ecologically grounded protocols with mandatory
subgroup performance reporting. For \textbf{regulators}, the actionable
step is to require evidence of ecologically valid evaluation as a
precondition for high-risk deployment authorization, rather than treating
post-market adverse event reporting as a sufficient safeguard. For
\textbf{health system leaders}, the lesson is that governance capacity
must be built through implementation support, not assumed as a
precondition.

The most concrete direction for future work is the empirical validation
of the ecological-validity-to-pilot-trap directional pathway: a multi-site
prospective study that tracks AI systems from regulatory authorization
through routine deployment, measuring how ecological validity at
authorization aligns with pilot-to-routine transition success. Such a
study would convert the documented pattern into a testable empirical
claim, and
would generate the evidence base that current regulatory frameworks
lack. More broadly, the field needs comparative empirical work on
non-Western implementation contexts to test whether the three-gap
framework travels across health system architectures, or whether it is
a framework specific to high-resource, English-dominant settings.

Ultimately, safe and sustainable clinical AI requires an
\textbf{integrated governance-and-evaluation ecosystem} in which technical
validation, organizational capacity-building, regulatory oversight, and
participatory accountability evolve together through implementation ---
not as independent sequential phases. Closing the ecological validity
gap is the most tractable point of intervention, since it conditions
both the evidence on which governance operates and the implementation
trajectories that governance must oversee.
