% =============================================================================
% 07_ACKNOWLEDGEMENTS
% =============================================================================

\section*{Acknowledgements}
\label{sec:acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

The author acknowledges with gratitude the contributions of the entire research
community whose published work made this systematic review possible, and
extends particular appreciation to the authors of the four cornerstone papers
analyzed in depth — Pham (2025), Tian et al. (2026), Hussein et al. (2026),
and Palama et al. (2026) — for their foundational contributions to the field.
The automated daily collection and analysis pipeline that supported this review
was built using open-source tools and freely available APIs; the author
acknowledges the maintainers of the underlying software dependencies.

The development of this literature review was supported by the LiGHT program
at EPFL. The author thanks supervisors Sahaj Vaidya and Mary-Anne Hartley for
their guidance and substantive feedback throughout the research process, and
acknowledges the contributions of the broader LiGHT writing community for
providing the scientific writing framework that informed the structure of
this report.

The pipeline that informed the corpus construction for this review made use of
freely available bibliographic APIs provided by PubMed, arXiv, Semantic Scholar,
and DBLP. The author acknowledges these services and their host institutions.

This research received no specific grant funding. The author declares no
conflicts of interest relevant to this work.