% =============================================================================
% 07_APPENDIX — Supplementary Materials
% =============================================================================

\appendix
\section*{Appendix}
\label{sec:appendix}
\addcontentsline{toc}{section}{Appendix}

% =============================================================================
% A.1 PRISMA FLOW DIAGRAM
% =============================================================================

\subsection{PRISMA Flow Diagram}
\label{sec:prisma}

% [TODO: Insert PRISMA flow diagram here — text description provided]
The PRISMA flow diagram for this systematic review is described below.
In the final submission, this should be replaced with the official
PRISMA diagram graphic generated through the PRISMA flow diagram
tool (http://prisma.thetacollaborative.ca/).

\begin{itemize}
\item \textbf{Records identified through database search (n = 246):}
  PubMed: [n1]; Semantic Scholar: [n2]; arXiv: [n3]; Google Scholar: [n4].
\item \textbf{Records after duplicates removed (n = [N]):} Merged and deduplicated.
\item \textbf{Records screened at abstract level (n = [N]):}
  Excluded: not health AI (n=x); not peer-reviewed (n=y); outside date range (n=z).
\item \textbf{Full-text assessed for eligibility (n = [N]):}
  Excluded at full-text: no evaluation/governance focus (n=x);
  predatory journal (n=y); other reason (n=z).
\item \textbf{Studies included in final corpus (n = [N]):}
  Empirical clinical studies: [n1]; Conceptual/policy analyses: [n2];
  Systematic/scoping reviews: [n3].
\end{itemize}

% =============================================================================
% A.2 ANNOTATED BIBLIOGRAPHY — top 10 selected papers
% =============================================================================

\subsection{Annotated Bibliography: Top 10 Selected Papers}
\label{sec:annotated_bib}

% [TODO: This can go in the appendix or as a separate document]
% Brief annotations (3–5 sentences) for each of the top 10 papers,
% explaining relevance to each of the 3 research sub-questions.

\medskip
\noindent\textbf{\underline{Paper 1 — Pham (2025): Ethical and legal considerations in healthcare AI}}
\textit{Relevance:} Covers all 5 themes. Serves as the comprehensive intro
paper for the review — bridges evaluation, governance, ethics, and implementation.

\medskip
\noindent\textbf{\underline{Paper 2 — Tian et al. (2026): From Pilot Trap to Institutional Capacity}}
\textit{Relevance:} Core paper for Sub-Question II (governance structures) and
Sub-Question III (implementation barriers). Provides the empirical foundation
for the "pilot trap" concept and the 6-module governance framework.

\medskip
\noindent\textbf{\underline{Paper 3 — OpenMRF (2026): Open-source MRF framework}}
\textit{Relevance:} Sub-Question I (evaluation methods) — demonstrates
reproducibility and open-science evaluation approaches in imaging AI.
\textbf{Note:} Verify whether this paper has a real peer-reviewed citation
before final submission; it may currently exist only as an arXiv preprint.

\medskip
\noindent\textbf{\underline{Paper 4 — Bailes et al. (2026): TRAC-Pain Protocol}}
\textit{Relevance:} Sub-Question I (evaluation methods) — methodological
template for prospective, longitudinal, ecologically grounded evaluation.
Demonstrates how evaluation methodology can be designed to capture
real-world clinical complexity.

\medskip
\noindent\textbf{\underline{Paper 5 — Warraich, Tazbaz \& Califf (2024): FDA Perspective}}
\textit{Relevance:} Sub-Question II (regulatory governance) — authoritative
regulatory overview. Essential reading for the regulatory landscape
(FDA and international coordination) section.

\medskip
\noindent\textbf{\underline{Paper 6 — Bouderhem (2025): Shaping the future of AI in healthcare through ethics and governance}}
\textit{Relevance:} Sub-Question II (participatory governance) — ethics-first
approach emphasizing community engagement and accountability mechanisms.

\medskip
\noindent\textbf{\underline{Paper 7 — Abdul (2025): AI-Driven Ethical Governance}}
\textit{Relevance:} Sub-Question II (adaptive governance) — balance between
innovation and accountability across data-intensive sectors including healthcare.
Useful for comparative sectoral perspective.

\medskip
\noindent\textbf{\underline{Paper 8 — Kéri (2026): Simulated consciousness in psychiatry AI}}
\textit{Relevance:} Sub-Question III (implementation barriers, safety risks) and
Sub-Question II (governance) — unique contribution on psychiatric AI failure
modes (simulated empathy, illusion of attachment) not addressed elsewhere.

\medskip
\noindent\textbf{\underline{Paper 9 — Vista4D (2026): Video reshooting with 4D point clouds}}
\textit{Relevance:} \textbf{EXCLUDED} from final corpus. This is a computer
vision framework for video manipulation, not a health AI paper. Tagged in error
by the pipeline. Do not include in final review.

\medskip
\noindent\textbf{\underline{Paper 10 — Abulibdeh, Celi \& Sejdić (2025): Illusion of safety — FDA report}}
\textit{Relevance:} Sub-Question I (evaluation — post-market monitoring gap)
and Sub-Question II (regulatory governance) — direct critique of FDA approval
process, calls for adaptive community-engaged regulation. Essential counterpoint
to Warraich et al.'s FDA perspective.

% =============================================================================
% A.3 KEYWORD THEMES — full keyword lists
% =============================================================================

\subsection{Search Keyword Themes}
\label{sec:keyword_themes}

% [TODO: Insert the full keyword theme table for transparency]
See Table~\ref{tab:keywords} for a full listing of keywords used across
the five thematic axes for the systematic search.

\begin{table}[h]
\centering
\begin{threeparttable}
\caption{Search keyword thematic axes and associated terms}
\label{tab:keywords}
\begin{tabularx}{\linewidth}{p{3.5cm} X}
\toprule
\textbf{Thematic Axis} & \textbf{Keywords} \\
\midrule
AI Evaluation \& Validation &
AI evaluation, AI validation, AI auditing, algorithmic auditing,
model evaluation, AI testing, AI assessment, external validation,
real-world validation, deployment evaluation, post-deployment monitoring,
continuous evaluation, AI performance assessment, bias detection,
robustness testing, AI reliability, explainability, safety evaluation \\
\midrule
Participatory Governance &
participatory governance, participatory democracy, community participation,
stakeholder engagement, citizen participation, deliberative democracy,
co-governance, collaborative governance, participatory AI, democratic AI,
community-centered AI, public interest technology, algorithmic accountability \\
\midrule
Adaptive Regulation &
adaptive governance, adaptive regulation, agile regulation, responsive
regulation, dynamic regulation, iterative regulation, experimental
regulation, learning regulation, regulatory sandboxes, living labs,
evidence-based regulation, outcome-based regulation, AI governance,
AI oversight, continuous compliance \\
\midrule
Evidence \& Implementation &
evidence-based policy, knowledge translation, research utilization,
evidence synthesis, implementation science, knowledge mobilization,
actionable evidence, evidence gaps, implementation barriers,
policy uptake, research impact, evidence quality \\
\midrule
Clinical Health AI &
clinical AI, health AI, medical AI, clinical decision support,
diagnostic AI, predictive modeling in healthcare, digital health,
AI safety in healthcare, clinical validation \\
\bottomrule
\end{tabularx}
\end{threeparttable}
\end{table}

% =============================================================================
% A.4 QUALITY APPRAISAL CRITERIA
% =============================================================================

\subsection{Quality Appraisal Rubric}
\label{sec:quality_rubric}

% [TODO: Insert quality rubric table if needed]
% For transparency, the bespoke quality rubric applied to conceptual and
% policy papers is available in full in the supplementary materials.
% Key dimensions: (1) explicit research question, (2) transparent methods,
% (3) balanced evidence, (4) coherent argumentation.

\end{document}